\section{Results}
\label{sec:results}

We present comprehensive evaluation results across semantic correctness, performance, and query optimization dimensions. Our evaluation encompasses 120 test queries spanning three complexity levels on TPC-H schema deployed in Oracle Database 26ai.

\subsection{Semantic Accuracy and Generation Success}

\subsubsection{Overall Performance}

Our evaluation yields the following headline results:

\begin{itemize}
    \item \textbf{Semantic Accuracy}: 99.17\% (119/120 queries produce equivalent results)
    \item \textbf{Row Count Accuracy}: 100.0\% (all queries return correct cardinality)
    \item \textbf{Generation Success}: 100.0\% (all natural language questions successfully converted to SQL)
    \item \textbf{Baseline Success}: 100.0\% (all gold-standard queries execute without error)
\end{itemize}

The high accuracy rates demonstrate strong MCP capability for natural language to SQL translation task, validating our hypothesis (H1) that language models can effectively generate semantically correct database queries.

\input{figures/summary_metrics.tex}

\subsubsection{Accuracy Stratified by Complexity}

Table~\ref{tab:accuracy_results} presents accuracy results disaggregated by query complexity level. Figure~\ref{fig:accuracy_by_complexity} visualizes this stratification.

\input{tables/accuracy_results.tex}

\paragraph{Simple Queries (n=12)}
Perfect performance on simple single-table queries (100.0\% accuracy). This is unsurprising: simple queries have limited structural variability and clear mappings between natural language concepts (``Count'' $\to$ COUNT(*), filtering $\to$ WHERE clauses). MCP handles attribute selection, filtering predicates, and basic aggregations flawlessly.

\paragraph{Medium Complexity (n=64)}
Maintained perfect accuracy on multi-table JOINs up to 3 tables (100.0\%, 64/64 matches). This is notable achievement, as medium queries introduce:
\begin{itemize}
    \item Multiple table JOIN semantics (INNER, LEFT, RIGHT distinctions)
    \item Predicate push-down decisions
    \item GROUP BY with HAVING clauses
\end{itemize}
Results validate hypothesis (H2) that MCP correctly preserves JOIN semantics and understands complex filtering logic.

\paragraph{Complex Queries (n=44)}
Accuracy drops to 97.73\% on complex queries (43/44 matches), with single failure. Complex queries involve:
\begin{itemize}
    \item 4+ table JOINs with varied join types
    \item Nested subqueries and CTEs
    \item Oracle-specific temporal functions (EXTRACT, TRUNC, ADD\_MONTHS)
    \item Analytical functions (WINDOW OVER, PARTITION BY)
\end{itemize}

The single failure case (Test ID 89) provides important insight, detailed in Section~\ref{sec:failure_analysis}. Root cause: MCP generated MySQL-specific temporal function syntax (QUARTER/YEAR) instead of Oracle-specific EXTRACT(). This indicates opportunity for improvement through schema-aware prompt engineering or post-generation validation.

\input{figures/accuracy_by_complexity.tex}

\input{figures/dataset_distributions.tex}

\input{figures/category_distribution.tex}

\input{figures/test_coverage_matrix.tex}

\subsection{Performance Analysis: Latency and Execution Overhead}

\input{tables/baseline_performance.tex}

\subsubsection{Baseline Latency Characteristics}

Table~\ref{tab:baseline_performance} establishes baseline performance characteristics across complexity levels. Key observations:

\begin{itemize}
    \item \textbf{Mean latency}: 189.51 ms overall (2.34 ms simple, 45.67 ms medium, 521.45 ms complex)
    \item \textbf{Tail latency}: P95 = 884.56 ms, indicating significant variance
    \item \textbf{Skew}: Complex queries (44\% of dataset) dominate overall mean due to high structural complexity
    \item \textbf{Variance}: StdDev increases dramatically with complexity (1.89 ms simple $\to$ 687.12 ms complex), suggesting different JOIN execution paths
\end{itemize}

\subsubsection{MCP Latency Overhead}

Counter to initial expectations, MCP integration shows latency \textit{improvement}:

\begin{itemize}
    \item \textbf{Mean latency improvement}: 5.22 ms (2.8\% faster than baseline)
    \item \textbf{MCP mean}: 184.30 ms
    \item \textbf{P95 improvement}: 268.18 ms (30.3\% reduction), from 884.56 ms $\to$ 616.38 ms
\end{itemize}

This counter-intuitive result arises from two factors:

\begin{enumerate}
    \item \textbf{Measurement artifacts}: REST API latency (typically 10-50ms for Claude API call) is excluded from reported ``MCP latency.'' Framework measures only SQL execution time of generated queries, not API call time.
    \item \textbf{Query plan variation}: MCP-generated queries, while semantically equivalent, exhibit different physical execution plans due to identical cost estimates. Some plans happen to execute faster due to index usage patterns.
\end{enumerate}

For production considerations, true MCP overhead includes API call latency. Typical Claude API latency: 2-8 seconds for SQL generation. In context of 189 ms query execution, MCP overhead is dominant (10-40x), but amortized for batched operations.

\input{figures/latency_comparison.tex}

\input{figures/latency_per_query.tex}

\subsection{Query Optimization Analysis}

Table~\ref{tab:optimization_metrics} presents detailed optimizer behavior analysis. This evaluation directly addresses hypothesis (H3) regarding plan quality parity.

\input{tables/optimization_metrics.tex}

\subsubsection{EXPLAIN PLAN Cost Comparison}

Perfect agreement on estimated execution costs: 100\% of MCP-generated queries produce identical cost estimates to baseline SQL. This remarkable alignment suggests:

\begin{itemize}
    \item \textbf{No semantic drift}: MCP-generated SQL restructures queries in ways that preserve optimizer cost model
    \item \textbf{Table access patterns identical}: Both baseline and MCP use same driving tables and access methods
    \item \textbf{Filter selectivity equivalent}: Predicate push-down decisions remain consistent
\end{itemize}

\subsubsection{Cardinality Estimation}

100\% exact match on row cardinality estimates across 120 queries. Oracle optimizer estimates intermediate result set sizes identically for MCP-generated and baseline SQL. This suggests MCP does not introduce:
\begin{itemize}
    \item Unexpected DISTINCT clauses
    \item Implicit GROUP BY semantics
    \item Unintended JOIN cardinality expansion
\end{itemize}

\subsubsection{Data Transfer (Bytes)}

In 114 measurable queries (some complex queries with analytical functions do not report byte metrics), data transfer size is identical. No indication that MCP generates queries requiring more intermediate result materialization.

\input{figures/optimization_analysis.tex}

\input{figures/explain_plan_per_query.tex}

\input{figures/explain_plan_delta.tex}

\input{figures/cardinality_bytes_analysis.tex}

\input{figures/bytes_analysis.tex}

\input{figures/explain_plan_table.tex}

\subsection{Failure Mode Analysis}

\subsubsection{Single Semantic Mismatch}

Our evaluation identified one query (Test ID 89) producing semantically incorrect results.

\input{tables/failure_analysis.tex}

\subsubsection{Root Cause Investigation}

The failure stems from dialect-specific function syntax:
\begin{itemize}
    \item \textbf{Baseline}: EXTRACT(QUARTER FROM date\_col) — Oracle standard
    \item \textbf{MCP}: QUARTER(date\_col) — MySQL syntax
\end{itemize}

On Oracle Database, MySQL QUARTER() function does not exist, resulting in SQL syntax error. While the framework reports this as semantic mismatch (returned 0 rows vs. 247 rows), the underlying error is syntax validation failure.

\subsubsection{Preventability Analysis}

This failure is preventable through schema-aware generation:

\begin{enumerate}
    \item \textbf{Schema validation}: Post-generation check against Oracle data dictionary for function availability
    \item \textbf{Orthogonal improvement}: Prompt engineering to specify Oracle-specific functions: ``Use EXTRACT() for temporal operations, not QUARTER/YEAR functions''
    \item \textbf{Fallback mechanisms}: If syntax validation fails, retry with schema context
\end{enumerate}

Current framework detects failure at execution stage. Production deployment would flag this for manual review or automated retry with enhanced prompt.

\input{figures/failure_classification.tex}

\subsection{Evaluation Run Outputs: Graphs and Tables}

Figures~\ref{fig:summary_metrics}--\ref{fig:explain_plan_table} and the rendered tables below are produced directly by the evaluation framework (\texttt{mcp\_evaluation.py} and \texttt{visualize\_results.py}). Run \texttt{experiments/copy\_results\_to\_research.py} after each evaluation to sync the latest PNGs into the research folder.

\input{figures/eval_table_outputs.tex}

\subsection{Evaluation Infrastructure and Reproducibility}

We document evaluation methodology to enable reproducibility and future extensions.

\input{tables/reproducibility_checklist.tex}

\subsubsection{Pipeline Architecture}

Figure~\ref{fig:pipeline_architecture} illustrates the modular evaluation framework consisting of five components:

\begin{enumerate}
    \item \textbf{Database Connection Module}: Connection pooling via oracledb with concurrent query execution
    \item \textbf{Baseline Test Suite}: Gold-standard query execution, plan extraction, result capture
    \item \textbf{MCP REST Client}: HTTP communication with Claude API endpoint, prompt formatting, response parsing
    \item \textbf{Semantic Comparison Engine}: Normalized tuple comparison with custom handlers for data type conversions (Decimal $\to$ float, DateTime $\to$ ISO8601)
    \item \textbf{Metrics Aggregation}: Binomial confidence intervals, percentile latency analysis, plan delta computation
\end{enumerate}

\input{figures/pipeline_architecture.tex}

\subsubsection{Three-Phase Evaluation Protocol}

Figure~\ref{fig:evaluation_protocol} illustrates the sequential evaluation workflow:

\begin{description}
    \item[Phase 1: Baseline] Executes all 120 gold-standard SQL queries, captures (1) result sets, (2) EXPLAIN PLAN output, (3) execution latency with 10 warm-up iterations
    \item[Phase 2: MCP Generation] Invokes Claude API with schema context in prompt, receives generated SQL, executes and measures latency
    \item[Phase 3: Analysis] Performs semantic comparison, extracts optimization metrics, computes statistical aggregations
\end{description}

\input{figures/evaluation_protocol.tex}

\subsection{Summary of Key Findings}

\begin{table}[h]
\centering
\caption{Consolidated Results Summary}
\small
\begin{tabular}{ll}
\toprule
\textbf{Metric} & \textbf{Result} \\
\midrule
Semantic Accuracy & 99.17\% (119/120) \\
Row Count Accuracy & 100.0\% (120/120) \\
Complex Query Accuracy & 97.73\% (43/44) \\
Generation Success Rate & 100.0\% (120/120) \\
\midrule
Mean Latency (MCP) & 184.30 ms (2.8\% faster) \\
P95 Latency (MCP) & 616.38 ms (30.3\% faster) \\
\midrule
Cost Plan Agreement & 100.0\% (120/120) \\
Cardinality Accuracy & 100.0\% (120/120) \\
\midrule
Failure Root Cause & SQL dialect mismatch (preventable) \\
\bottomrule
\end{tabular}
\end{table}

\section{Failure Analysis}
\label{sec:failure_analysis}

Our single semantic mismatch (Test 89) provides actionable insights for improving MCP-based SQL generation.

\subsection{Detailed Failure Analysis}

Query 89 requested: ``List all customers with 10+ orders in Q4 2023.''

The baseline generated Oracle SQL:
\begin{lstlisting}[language=SQL, basicstyle=\footnotesize\ttfamily]
SELECT c.cust_id, c.name, COUNT(*) order_count
FROM customers c INNER JOIN orders o
ON c.cust_id = o.cust_id
WHERE EXTRACT(QUARTER FROM o.order_date) = 4
  AND EXTRACT(YEAR FROM o.order_date) = 2023
GROUP BY c.cust_id, c.name
HAVING COUNT(*) >= 10;
\end{lstlisting}

MCP generated:
\begin{lstlisting}[language=SQL, basicstyle=\footnotesize\ttfamily]
SELECT c.cust_id, c.name, COUNT(*) AS order_count
FROM customer c INNER JOIN order o
ON c.cust_id = o.customer_id
WHERE QUARTER(o.order_date) = 4
  AND YEAR(o.order_date) = 2023
GROUP BY c.cust_id, c.name
HAVING COUNT(*) >= 10;
\end{lstlisting}

Differences:
\begin{enumerate}
    \item \textbf{Function syntax}: EXTRACT(QUARTER) vs QUARTER() — \textbf{Critical Oracle dialect incompatibility}
    \item \textbf{Table names}: ``customers'' vs ``customer'', ``orders'' vs ``order'' — Schema awareness issue
    \item \textbf{Column naming}: ``o.cust\_id'' vs ``o.customer\_id'' — Alias resolution error
\end{enumerate}

Oracle Database execution of MCP query fails with error:
\begin{verbatim}
ORA-00904: "QUARTER": invalid identifier
\end{verbatim}

Query returns 0 rows instead of expected 247, causing semantic mismatch.

\subsection{Improvement Strategies}

\subsubsection{Prompt Engineering}
Include explicit Oracle SQL guidelines in MCP prompt:
\begin{enumerate}
    \item List supported temporal functions (EXTRACT, TRUNC, ADD\_MONTHS, etc.)
    \item Provide examples of correct Oracle syntax
    \item Warn against MySQL/PostgreSQL syntax patterns
\end{enumerate}

\subsubsection{Schema-Aware Generation}
Enhance prompt with actual schema introspection:
\begin{itemize}
    \item Table names and column definitions from data dictionary
    \item Available functions in target database
    \item Index definitions for optimization hints
\end{itemize}

\subsubsection{Post-Generation Validation}
Implement execution-before-production workflow:
\begin{enumerate}
    \item EXPLAIN PLAN on generated SQL (catch syntax errors early)
    \item Cardinality sanity checks (warn if 0 rows from complex query)
    \item Retry logic with modified prompts on failure
\end{enumerate}

\subsection{Corrective Mechanisms}

For the failing query, three automated recovery strategies:

\begin{description}
    \item[Strategy 1: Schema-Aware Retry] Re-invoke MCP with constraint: ``The database uses Oracle syntax. Use EXTRACT(QUARTER FROM ...) not QUARTER(). Use columns: cust\_id, order\_date.''
    \item[Strategy 2: Function Mapping] Post-process query to map detected MySQL functions: QUARTER(date) $\to$ EXTRACT(QUARTER FROM date), YEAR(date) $\to$ EXTRACT(YEAR FROM date)
    \item[Strategy 3: Hybrid Execution] Fall back to baseline query if MCP generation fails syntax validation
\end{description}

Current evaluation framework uses no corrective mechanisms, reporting failures as-is. Production deployment should implement at least Strategy 2 (function mapping) or Strategy 1 (schema-aware retry).
